---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# htmed

```{r setup, message=FALSE}
library(htmed)
#library(survival)
library(dplyr) #
#library(ggraph) #
```

In this tutorial, we will use all the functions available in `htmed` package to illustrate an example of how high-throughput mediation analysis could be performed.

## Data

To do so, we will simulate some data. Imagine that we have 100 individuals for which we have normalised measures of 4 biochemical parameters: LDL, cholesterol, free cholesterol and triglycerides. We also have information about ventricle end-diastolic volume and end-systolic volume, in both right and left ventricles. In this case, the data is binary, with 0 for individuals with normal values, and 1 for individuals with ranges out of normality. Finally, we have heart oxygen consumption information, also normalised.

Our aim is to study the association between the biochemical parameters (_treatments_) and the heart oxygen consumption (_outcome_). In addition, we want to determine is any of the ventricle measures (_mediators_) somehow explains the underlying mechanism of the relationship between the treatments and the outcome.

```{r load_df}
set.seed(123)

n <- 100

# data for the mediators
df <- data.frame(
  LV.EDV = rbinom(n, 1, 0.5), 
  RV.EDV = rbinom(n, 1, 0.5), 
  LV.ESV = rbinom(n, 1, 0.5),  
  RV.ESV = rbinom(n, 1, 0.5)
)

# data for the treatment
df$LDL <- - 0.5 * df$LV.EDV + 0.3 * df$RV.EDV + 0.4 * df$LV.ESV + 0.6 * df$RV.ESV + rnorm(n)
df$cholesterol <- -0.6 * df$LV.EDV - 0.5 * df$RV.EDV - 0.4 * df$LV.ESV - 0.7 * df$RV.ESV + rnorm(n)
df$free.cholesterol <- + df$LV.ESV + rnorm(n)
df$triglycerides <- + 0.5 * df$RV.EDV + 0.8 * df$LV.ESV + rnorm(n)

# data for the outcome
df$oxyg_consump <- 1 - 1.2 * df$LV.EDV + 1.0 * df$LDL + 0.8 * df$cholesterol + 0.6 * df$free.cholesterol + 0.9 * df$triglycerides + rnorm(n)
df <- df %>% mutate(across(where(is.numeric), ~ as.numeric(scale(.))))
```

With this data, we hypothesize that some treatments could be responsible of the values for the outcome observed in the individuals, and that some mediators could explain the underlying mechanism of the relationship between the treatment and the outcome. 

Therefore, we are interested in testing the association between each treatment and outcome, through each mediator. In order to perform this mediation analyses, we first need to compute the fitted models for mediators and outcomes, respectively, something that can be done with the functions `mediator_models()` and `outome_models()`, respectively. However, before applying this function, we need to generate a new DataFrame in which we will specify what combination of treatment, mediator and outcome we want to test. The DataFrame (called `models` in this example), will be composed of as many rows as different analyses we want to perform and five different columns:

* outcome (character): the different outcomes that we are interested in predict. In this example we will work with just one outcome.
* treatment (character): the different treatments to test. 
* mediator (character): the different mediators to test.
* model.m.formula (character): it will inform about the formula that needs to be applied to compute the fitted models for mediator.
* model.y.formula (character): analog to model.m.formula but with the formulas for the fitted models for outcome.

As you can see, we will use the `data_models()` function to generate this DataFrame. By default, this function makes all possible combinations of the supplied vectors for outcome, treatment and mediator. If the user is interested in compute just some specific models, we recommend to build a DataFrame following the five-column structure of `models`.

```{r load_models}
# data("models_surv", package = "htmed")
outcome <- 'oxyg_consump'
treatment <- c('LDL', 'cholesterol', 'free.cholesterol', 'triglycerides')
mediator <- c('LV.EDV', 'RV.EDV', 'LV.ESV', 'RV.ESV')
    
models <- data_models(outcome = outcome, mediator = mediator, treatment = treatment)

```

## Simple usage
### Generatting models for mediators and outcomes

After generating `models`, we have all the necessary to apply the `mediator_models()` and `outome_models()` functions. By applying them, we will generate the fitted models for mediators and outcomes, for each combination of mediator, treatment and outcome, in an iterative manner. 
Both functions contain 6 arguments, being mandatory 5 of them:

* `column.models` (character): the name of the column that contains the model formulas in `data.models`
* `model.type` (function): the statistical analysis we are interested in applying over the data
* `data` (dataframe): the DataFrame with the values 
* `data.models` (dataframe): the DataFrame with the models to perform
* `model.name` (character): the name of the new column that will contain the fitted models for the mediator or outcome (depending on the function we are applying)

There are other 2 optional arguments: `outcome` will refer to the column from the `data.models` DataFrame that contains the information of the outcomes. Although this last argument is optional and does not need to be specified if we have just one outcome, it is highly recommended if you are dealing with more than one dependent variables.
The `data.split()` argument will be explained later.

Both functions also accept additional arguments (`...`) that are passed directly to the model function specified in `model.type`. This is useful when the chosen model requires extra parameters — for example, passing `family = binomial()` when using `glm` for logistic regression.
```{r generatting_models-1, message=FALSE}
# fitted models for the mediator
medANDout <- mediator_models(
    column.models='model.m.formula'
  , model.type=lm
  , data=df
  , data.models=models
  , model.name = 'model.M'
  ) 

# fitted models for the outcome
medANDout <- outcome_models(
    column.models='model.y.formula'
  , model.type=lm
  , data=df
  , data.models=medANDout
  , model.name = 'model.Y'
  ) 
```

The new dataframe will contain the same information than `models`, but with two additional columns:

* model.M: fitted models for mediators
* model.Y: fitted models for outcomes

### High-throughput mediation analysis

To apply high-throughput mediation, we execute the `htmed()` function over this data. From this point, we will just work with the DataFrame generated in the previous step `medANDout`.
`htmed()` requires the following arguments:

* data.models (dataframe): object with the fitted models for mediators and outcomes.
* column.modelm and column.modely (characters): these arguments refer to the columns from `data.models` that contain the fitted models for mediator and outcomes, respectively.
* treat, mediator and outcome (characters): these three arguments refer to the columns from `data.models` with the treatment, mediator and outcome information, respectively.
* data.split (character): will be explained later.

There is an additional argument `seed` that can be introduced to ensure reproducibility of results.

`htmed()` also accepts additional arguments (`...`) that are passed directly to `mediation::mediate()`. This is useful when the mediation model requires specific control options — for example, `control.value` and `treat.value` to define the reference levels of a non-binary treatment.
```{r applying_htmed-1, message=FALSE}
med_results <- htmed(
    data.models=medANDout
  , column.modelm = 'model.M'
  , column.modely = 'model.Y'
  , treat='treatment'
  , mediator='mediator'
  , outcome='outcome'
  )
```

```{r applying_htmed-2}
# med_results %>% View(.)

# Class of the mediation analysis results:
unlist(unique(lapply(med_results$oxyg_consump, function(x) {class(x)})))
```

`htmed()` will generate a list of lists in which the first level will be the different outcomes tested (one in this example), and the second level, the mediation analyses performed.

```{r applying_htmed-3}
# Outcomes tested
names(med_results)

# Analyses performed for oxyg_consump
names(med_results$oxyg_consump)
```

### Formatting the mediation results

If we explore the results, we will see that the output is in some way difficult to understand.
```{r formatting_data-1}
# med_results$oxyg_consump['`LV.EDV` ~ `cholesterol`'] 
```

We need to transform this data to simplify and make it more user-friendly for the visualizations that will be performed later. Using `format_med()` with just `med_results` as input, we will generate a DataFrame with the essential columns needed for the visualizations.

`format_med()` has an `estimate` argument that controls which estimate is reported when the mediation model has a control/treated split. It accepts `"average"` (default), `"control"`, or `"treated"`. For models without a split, this argument is ignored.

This function also computes two different kinds of adjusted p-value: adj.pval.PropMed.by_outcome and adj.pval.PropMed.all. The difference between both results is that, while the first one adjusts by each of the outcomes, the second adjusted p-value takes into account all the analyses from all the outcomes. The usage of one or the other will depend on the question that the investigator wants to answer.

```{r formatting_data-2}
# formatting data
format_results <- format_med(mediation.list = med_results)
```

### Visualization

Finally, we will be interested in visualise our results. The package has two available functions for this purposse.

We can create a scatterplot for each outcome with the `visual_htmed()` function. Run in the default model, just with the mandatory arguments, we will need to specify:

* mediation_form (lists of lists): the object that contain the results.
* outcome (characters): the outcome the user is interested in visualising.

```{r visualizing-1}
visual_outcome1_nosig <- visual_htmed(mediation.form = format_results, outcome = 'oxyg_consump')
visual_outcome1_nosig
```
This scatterplot will represents, for oxyg_consump, the relationship between the treatments and mediators, being the size of the dot proportional to the proportion of mediation, and the color, the estimation of mediation.

Another visualization can be done with `graph_htmed()` function. In the default mode of this function, the graph will display the treatments as the internal nodes, and the mediator as the external ones. Similar to what `visual_htmed()` does, the width of the edges is proportional to the proportion of mediation, and the color, to the estimation of mediation.
```{r visualizing-2}
graph_outcome1_nosig <- graph_htmed(mediation.form = format_results, outcome = 'oxyg_consump')
graph_outcome1_nosig
```

You might have notice that we have not filtered out any mediation analyses in the representation, which means that all results, independently of the level of significance, are visualized. This is what occurs by default in both `visual_htmed()` and `graph_htmed()` functions.

Nevertheless, you can filter the data by specifying the level of significance (`pval`) and the column to apply it (`pval.column`). In the following chunks, we will restrict our representation to mediation analyses with an p-value <= 0.05.
```{r visualizing-3}
visual_outcome1_adj0.05 <- visual_htmed(
  mediation.form = format_results
  , outcome = 'oxyg_consump'
  , pval.column = "pval.PropMed"
  , pval = 0.05)
visual_outcome1_adj0.05
```


```{r visualizing-4}
graph_outcome1_adj0.05 <- graph_htmed(
  mediation.form = format_results
  , outcome = 'oxyg_consump'
  , pval.column = "pval.PropMed"
  , pval = 0.05)
graph_outcome1_adj0.05
```

Additionally, `grapg_htmed()` allows you to custom some visualization parameters such as the size of the node, the size of the node names and the end of the arrows, as it will be shown in the next chunk.
```{r visualizing-5}
graph_htmed(
    mediation.form = format_results
  , outcome = 'oxyg_consump'
  , pval.column = "pval.PropMed"
  , pval = 0.05
  , size_node = 1.5
  , size_name = 1.3
  , end_arrow = 4)
```


## Splitting the data

Now, we will include a new column to the original DataFrame, to classify individuals depending on their comorbidity. 
```{r split}
df <- df %>% 
  mutate(split = sample(x = c('HighBP', 'Diabetes'), size = 100, replace = TRUE))

df %>% dplyr::select(split) %>% table(.)
```

Under the previous hypothesis, we were perfoming the analyses with the individuals as a whole. Nevertheless, we now think that the underlying mechanisms that leads to the outcome could differ depending on the condition observed in the individuals. `htmed` package allows to take this into account by some additional arguments present in the functions available.

### Generating models for mediators and outcomes

Therefore, we will execute `mediator_models()` `outcome_models()` and  iteratively to generate both fitted outcomes for mediator and outcome, by adding `data.split` argument. This will indicate the column from `df` that has the information to split with.
```{r generatting_models.split-1}
# fitted models for the mediator
medANDout.split <- mediator_models(
    column.models='model.m.formula'
  , model.type=lm
  , data=df
  , data.models=models
  , model.name = 'model.M'
  , data.split = 'split'
  ) 

# fitted models for the outcome
medANDout.split <- outcome_models(
    column.models='model.y.formula'
  , model.type=lm
  , data=df
  , data.models=medANDout.split
  , model.name = 'model.Y'
  , data.split = 'split'
  ) 
```

Comparing this new DataFrame (`medANDout.split`) with the one in which split was not applied (`medANDout`), we will notice that the new one doubles the dimension of the second one. This is because the fitted models have been performed independently for each condition.
```{r generatting_models.split-2}
dim(medANDout.split)

dim(medANDout)
```

### High-throughput mediation analysis
To perform the mediation analyses, we will add the same argument (`data.split`).
```{r applying_htmed.split-1, message=FALSE}
med_results.split <- htmed(
    data.models=medANDout.split
  , column.modelm = 'model.M'
  , column.modely = 'model.Y'
  , treat='treatment'
  , mediator='mediator'
  , outcome='outcome'
  , data.split = 'split'
  )
```

This new object will have three levels of lists. The first will be the outcomes tested, whereas the second one will be the conditions used to split. The third will be the different mediation analyses.
```{r applying_htmed.split-2}
# Outcomes tested
names(med_results.split)

# Conditions
names(med_results.split$oxyg_consump)

# Analyses for Diabetes
names(med_results.split$oxyg_consump$Diabetes)
```

### Formatting the mediation results
When we are dealing with splitted data, we need to explicit it to the `format_med()` function to format the data right. This will be done by setting to TRUE the `split` argument.
```{r formatting_data.split-1}
format_results.split <- format_med(med_results.split, split = TRUE)
```

### Visualization
```{r visualizing.split-1}
visual_outcome1_nosig.split <- visual_htmed(
    mediation.form = format_results.split
  , outcome = 'oxyg_consump'
  , data.split = 'split')
visual_outcome1_nosig.split
```

In this case, two scatterplots will be displayed, differenciating between the analyses from the two initial conditions (Diabetes and High Blood Pressure).

```{r visualizing.split-2}
graph_outcome1_nosig.split <- graph_htmed(
    mediation.form = format_results.split
  , outcome = 'oxyg_consump'
  , size_node = 0.65
  , data.split = 'split')

library(patchwork)
graph_outcome1_nosig.split$Diabetes + graph_outcome1_nosig.split$HighBP
```
The same occurs when applying `graph_htmed()` ; two graph are generated, depending on the condition studied.
